Containers and R---the Rockerverse and beyond

### 2019-12-13 manuscript.Rmd

---
title: "Containers and R---the Rockerverse and beyond"
author:
  - name: Daniel Nüst
    affiliation: University of Münster
    address:
      - Institute for Geoinformatics
      - Heisenbergstr. 2
      - 48149 Münster, Germany
      - \orcid{0000-0002-0024-5046}
    email: daniel.nuest@uni-muenster.de
  - name: Carl Boettiger
    #affiliation: -
    #email: -
  - name: Robrecht Cannoodt
    affiliation: Ghent University
    address:
      - Data Mining and Modelling for Biomedicine group
      - VIB Center for Inflammation Research
      - Technologiepark 71
      - 9052 Ghent, Belgium
      - \orcid{0000-0003-3641-729X}
    email: robrecht@cannoodt.dev
  - name: Ellis Hughes
  - name: Dirk Eddelbuettel
    affiliation: University of Illinois at Urbana-Champaign
    address:
      - Department of Statistics
      - Illini Hall, 725 S Wright St
      - Champaign, IL 61820, USA
      - \orcid{0000-0001-6419-907X}
    email: dirkd@eddelbuettel.com
  - name: Mark Edmondson
    affiliation: IIH Nordic A/S, Google Developer Expert for GCP
    email: mark@markedmondson.me
  - name: Colin Fay
    affiliation: ThinkR
    address:
      - 5O rue Arthur Rimbaud
      - 93300 Aubervilliers, France
      - \orcid{0000-0001-7343-1846}
    email: contact@colinfay.me
  - name: Ben Marwick
    #affiliation: -
    #email: -
  - name: Karthik Ram
    affiliation: Berkeley Institute for Data Science
    address:
      - University of California
      - Berkeley, CA 94720, USA
      - \orcid{0000-0002-0233-1757}
    email: karthik.ram@berkeley.edu
  - name: Noam Ross
    affiliation: EcoHealth Alliance
    address:
      - 460 W 34th St., Ste. 1701
      - New York, NY 10001, USA
      - \orcid{0000-0002-0233-1757}
    email: ross@ecohealthalliance.org
  - name: Nan Xiao
    affiliation: Seven Bridges Genomics
    address:
      - 529 Main St, Suite 6610
      - Charlestown, MA 02129, USA
      - \orcid{0000-0002-0250-5673}
    email: me@nanx.me
  - name: Lori Shepherd
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
      - Elm & Carlton Streets
      - Buffalo, New York, 14263, USA
      - \orcid{0000-0002-5910-4010}
    email: lori.shepherd@roswellpark.org
  - name: Nitesh Turaga
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
    - Elm & Carlton Streets
    - Buffalo, New York, 14263, USA
    - \orcid{0000-0002-0224-9817}
    email: nitesh.turaga@roswellpark.org
abstract: >
  The Rocker project provides widely-used Docker images for R across different application scenarios.
  This articles surveys describe downstream projects building upon Rocker.
  We also look beyond Rocker to other projects connecting containerisation with R.
  These use cases and the diversity of applications demonstrate the power of Rocker and containerisation for collaboration, effectivity, scalability, and transparency.
# blank footnote? https://tex.stackexchange.com/questions/170511/footnotes-without-numbering
output:
  rticles::rjournal_article:
    keep_tex: true
    includes:
      in_header: header.tex
---

@@

**Microsoft R Open** ([MRO](https://github.com/nuest/mro-docker)) is an R distribution formerly known as Revolution R Open (RRO) before Revolution Analytics was acquired by Microsoft.
MRO is compatible with main R and it's packages.sss
"It includes additional capabilities for improved performance, reproducibility, and platform support." \citep{microsoft_mro}
Most notably these capabilities are the MRAN repository, which is enabled by default, and the (optional) integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread}.
MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images `nuest/mro` are available on Docker Hub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) and on GitHub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) respectively.
The images are inspired by the Rocker images and can be used much in the same fashion, effectively a drop-in replacement allowing users to quickly evaluate if the benefits of MRO + Intel® MKL apply to their use case.
Version-tagged images are provided for the latest bugfix release of recent R versions.
Extended license information about MKL is printed at every startup.

@@

- `RSelenium`
- [`googleComputeEngineR`](https://cloudyr.github.io/googleComputeEngineR/) (function `docker_run`)
  - I don't think `docker_run` is main focus for containers for this package, its more how it uses Docker to launch custom environments that can be templated (e.g. rstudio, shiny etc.) and for parrallisation e.g. `library(future)`  using GCE VMs as endpoints. [@MarkEdmondson1234]
- [`analogsea`](https://github.com/sckott/analogsea) (digital ocean R client)
- [`plumber`](https://www.rplumber.io/docs/hosting.html#docker)
- [production deployment of neural networks](https://github.com/tmobile/r-tensorflow-api) [@jnolis, @nolistic]

>>>>

---
title: "Containers and R---the Rockerverse and beyond"
author:
  - name: Daniel Nüst
    affiliation: University of Münster
    address:
      - Institute for Geoinformatics
      - Heisenbergstr. 2
      - 48149 Münster, Germany
      - \orcid{0000-0002-0024-5046}
    email: daniel.nuest@uni-muenster.de
  - name: Carl Boettiger
    #affiliation: -
    #email: -
  - name: Robrecht Cannoodt
    affiliation: Ghent University
    address:
      - Data Mining and Modelling for Biomedicine group
      - VIB Center for Inflammation Research
      - Technologiepark 71
      - 9052 Ghent, Belgium
      - \orcid{0000-0003-3641-729X}
    email: robrecht@cannoodt.dev
  - name: Ellis Hughes
  - name: Dirk Eddelbuettel
    affiliation: University of Illinois at Urbana-Champaign
    address:
      - Department of Statistics
      - Illini Hall, 725 S Wright St
      - Champaign, IL 61820, USA
      - \orcid{0000-0001-6419-907X}
    email: dirkd@eddelbuettel.com
  - name: Mark Edmondson
    affiliation: IIH Nordic A/S, Google Developer Expert for GCP
    email: mark@markedmondson.me
  - name: Colin Fay
    affiliation: ThinkR
    address:
      - 5O rue Arthur Rimbaud
      - 93300 Aubervilliers, France
      - \orcid{0000-0001-7343-1846}
    email: contact@colinfay.me
  - name: Ben Marwick
    #affiliation: -
    #email: -
  - name: Karthik Ram
    affiliation: Berkeley Institute for Data Science
    address:
      - University of California
      - Berkeley, CA 94720, USA
      - \orcid{0000-0002-0233-1757}
    email: karthik.ram@berkeley.edu
  - name: Noam Ross
    affiliation: EcoHealth Alliance
    address:
      - 460 W 34th St., Ste. 1701
      - New York, NY 10001, USA
      - \orcid{0000-0002-0233-1757}
    email: ross@ecohealthalliance.org
  - name: Nan Xiao
    affiliation: Seven Bridges Genomics
    address:
      - 529 Main St, Suite 6610
      - Charlestown, MA 02129, USA
      - \orcid{0000-0002-0250-5673}
    email: me@nanx.me
  - name: Lori Shepherd
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
      - Elm & Carlton Streets
      - Buffalo, New York, 14263, USA
      - \orcid{0000-0002-5910-4010}
    email: lori.shepherd@roswellpark.org
  - name: Nitesh Turaga
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
    - Elm & Carlton Streets
    - Buffalo, New York, 14263, USA
    - \orcid{0000-0002-0224-9817}
    email: nitesh.turaga@roswellpark.org
  - name: Hong Ooi
    affiliation: Microsoft
    address:
    - Level 5, 4 Freshwater Place
    - Southbank, VIC 3006, Australia
    email: hongooi@microsoft.com
abstract: >
  The Rocker project provides widely-used Docker images for R across different application scenarios.
  This articles surveys describe downstream projects building upon Rocker.
  We also look beyond Rocker to other projects connecting containerisation with R.
  These use cases and the diversity of applications demonstrate the power of Rocker and containerisation for collaboration, effectivity, scalability, and transparency.
# blank footnote? https://tex.stackexchange.com/questions/170511/footnotes-without-numbering
output:
  rticles::rjournal_article:
    keep_tex: true
    includes:
      in_header: header.tex
---

@@

**Microsoft R Open** ([MRO](https://github.com/nuest/mro-docker)) is an R distribution formerly known as Revolution R Open (RRO) before Revolution Analytics was acquired by Microsoft.
MRO is compatible with main R and its packages.
"It includes additional capabilities for improved performance, reproducibility, and platform support." \citep{microsoft_mro}
Most notably these capabilities are the MRAN repository, which is enabled by default, and the (optional) integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread}.
MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images `nuest/mro` are available on Docker Hub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) and on GitHub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) respectively.
The images are inspired by the Rocker images and can be used much in the same fashion, effectively a drop-in replacement allowing users to quickly evaluate if the benefits of MRO + Intel® MKL apply to their use case.
Version-tagged images are provided for the latest bugfix release of recent R versions.
Extended license information about MKL is printed at every startup.

@@

- `RSelenium`
- [`googleComputeEngineR`](https://cloudyr.github.io/googleComputeEngineR/) (function `docker_run`)
  - I don't think `docker_run` is main focus for containers for this package, its more how it uses Docker to launch custom environments that can be templated (e.g. rstudio, shiny etc.) and for parrallisation e.g. `library(future)`  using GCE VMs as endpoints. [@MarkEdmondson1234]
- [`analogsea`](https://github.com/sckott/analogsea) (digital ocean R client)
- [`AzureContainers`](https://cran.r-project.org/package=AzureContainers): Umbrella package for working with containers in Microsoft's Azure cloud. Provides interfaces to three Azure services: Container Registry, Container Instances (for running individual containers), and Kubernetes Services (for orchestrated deployments)

## Using R to power enterprise software in production environments [@jnolis, @nolistic]

R has been historically viewed as a tool for analysis and scientific research--not for creating software that corporations can rely on to continously run in real time. However, thanks to advancements in R running as a web service, along with with the ability to deploy R in docker containers, modern enterprises are now capable of having real-time machine learning powered by R.

The backbone of this work is the _[plumber](https://www.rplumber.io/docs/hosting.html#docker)_ package, created by Jeff Allen. The plumber package allows R to be used to host web services through RESTful APIs. When R is running on a server, a different server can ask R to make a computation through an HTTP request and get an immediate response. For instance, if a data scientist at a company creates a machine learning model to tell which customers are likely to make a repeat purchase, if they serve the model through a plumber API then someone else on the marketing team could write software that sends different emails depending on that real-time prediction. Plumber syntax is simple to read and understand: the API is created by commenting above function names in an R script.

```
#* Return the sum of two numbers
#* @get /random
function(){
  runif()
}
```
_Example of a plumber API to generated a random number by visiting `http://127.0.0.1/random`_

Plumber becomes far more powerful with the addition of Docker. Docker can be used to create images that [include both R and plumber](https://www.rplumber.io/docs/hosting.html#docker). Then when the image is deployed, that creates the web endpoints that other services can hit and call the R code. This pattern of deploying code mimics those used by software engineering services created in Java, Python, and other modern languages. By using Docker to deploy plumber APIs, R can be used alongside those langages as a first class member of a software engineering technical stack.

At T-Mobile, R is used to power real-time services that are called over a million times a day. T-Mobile created a set of neural network machine learning natural language processing models to help customer care agents manage text-based messages for customers. The models are convolutional neural networks that use the RStudio Keras package and are built on top of a Rocker Docker image. Since the models power tools for agents and customers, they need to have extremely high uptime and reliability--and they found that R is able to perform. The AI @ T-Mobile team open sourced [their docker images](https://github.com/tmobile/r-tensorflow-api  which are available on GitHub.

## Deployment to the cloud [@hongooi]

While Plumber provides the low-level infrastructure for turning a statistical model into a predictive service, for production purposes it is usually necessary to take into account considerations such as scalability, reliability and ease of management. A single container is limited in the volume of requests it can service. Furthermore, if the container goes down, or is occupied with processing a computationally intensive task, the service becomes unavailable until it is restarted.

These issues can be mitigated by deploying not a single container, but a _cluster_ of containers, orchestrated as a single deployment. The deployment system of choice at the time of writing is [Kubernetes](https://kubernetes.io). This allows you to deploy a single image to a pool of servers, which can then serve requests in parallel. Client-side requests are made to the cluster ingress endpoint, which then redirects them to individual servers. The orchestrator can identify cluster nodes that have failed, and restart them automatically; it can also detect when there is no capacity to service more requests, and start additional servers and containers as needed.

While these load-balancing and autoscaling properties of a Kubernetes cluster are invaluable, they are still limited by the hardware resources that it has access to. A cluster cannot scale out if it has run out of servers. This issue in turn can be addressed by deploying not to on-premise servers but to the cloud, where effectively unlimited computing resources are available (as long as one can pay for them). From a management perspective, cloud deployments have the additional advantage that hardware maintenance becomes the responsibility of the cloud provider, eliminating another pain point. Orchestrator services are available from all the major cloud service providers, including Amazon Web Services (AWS), Google Cloud Platform, and Microsoft Azure.

An example of an R package for working with containers in the cloud is AzureContainers, which interacts with three Azure services: Container Instances (for running a single container), Container Registry (a private docker registry service), and Kubernetes Service (for Kubernetes on Azure). AzureContainers provides a lightweight yet powerful interface for working with Resource Manager, the framework for deploying and managing arbitrary resources in Azure. On the client side, it provides simple shells to the Docker, Kubectl and Helm tools that are commonly used for managing containers and Kubernetes clusters.

Here is some simple code that pushes a Docker image to an Azure Container Registry, and then creates a deployment using Azure Kubernetes Service. This is a slightly modified version of the code from the "Deploying a prediction service with Plumber" AzureContainers vignette. The code makes use of a yaml file to define the deployment and predictive service, which is standard practice with Kubernetes; the yaml can be seen in the aforementioned vignette.

```r
library(AzureContainers)
# create a resource group for our deployments
deployresgrp <- AzureRMR::get_azure_login()$
    get_subscription("subscription_id")$
    create_resource_group("deployresgrp", location="australiaeast")

### create a container registry
deployreg_svc <- deployresgrp$create_acr("deployreg")

# build our image (a random forest on the Boston housing data, available in the MASS package)
call_docker("build -t bos-rf .")

# upload the image to Azure
deployreg <- deployreg_svc$get_docker_registry()
deployreg$push("bos-rf")

### create a Kubernetes cluster with 2 nodes
deployclus_svc <- deployresgrp$create_aks("deployclus", agent_pools=aks_pools("pool1", 2))

# grant the cluster pull access to the registry
gr <- AzureGraph::get_graph_login()
aks_app_id <- deployclus$properties$servicePrincipalProfile$clientID
reg$add_role_assignment(gr$get_app(aks_app_id), "Acrpull")

# get the cluster endpoint
deployclus <- deployclus_svc$get_cluster()

# create and start the service
deployclus$create("bos-rf.yaml")
```

Once the service has been created, we can check on its status and obtain predictions:

```r
deployclus$get("deployment bos-rf")
#> Kubernetes operation: get deployment bos-rf  --kubeconfig=".../kubeconfigxxxx"
#> NAME      DESIRED   CURRENT   UP-TO-DATE   AVAILABLE   AGE
#> bos-rf    1         1         1            1           5m

deployclus$get("service bos-rf-svc")
#> Kubernetes operation: get service bos-rf-svc  --kubeconfig=".../kubeconfigxxxx"
#> NAME         TYPE           CLUSTER-IP   EXTERNAL-IP     PORT(S)          AGE
#> bos-rf-svc   LoadBalancer   10.0.8.189   xxx.xxx.xxx.xxx 8000:32276/TCP   5m

response <- httr::POST("http://xxx.xxx.xxx.xxx:8000/score",
    body=list(df=MASS::Boston[1:10,]), encode="json")
httr::content(response, simplifyVector=TRUE)
#> [1] 25.9269 22.0636 34.1876 33.7737 34.8081 27.6394 21.8007 22.3577 16.7812 18.9785
```

Note that Plumber, by itself, provides no authentication or security features; anybody who knows the cluster's IP address can access the predictive service. At minimum, a basic authentication layer should be added to any Kubernetes deployment. Standard Kubernetes procedure involves creating an ingress controller, for example with the nginx or traefik reverse proxy software, and one can then add application-specific authentication layers on top of that (for example, to authenticate with a user's Azure Active Directory or LDAP credentials). More details can be found from most Kubernetes learning resources, and are outside the scope of this paper.
